Behavioral Analysis of the Agent-Based Community Grid Solution for the Large Hadron Collider beauty Experiment

### 2011-11-13 paper_update.tex
halfway there
Chapter 3 finally finished!

\usepackage{color}
\usepackage{url}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{listings}
\definecolor{light-gray}{gray}{0.80}
%\usepackage[center]{caption}
\usepackage{authblk}
\begin{document}
\title{Behavioral Analysis of the Agent-Based Community Grid Solution for the Large Hadron Collider beauty Experiment}
\author[1,3]{Daniela Remenska}
\author[2]{Tim A.C. Willemse}
\author[1]{Henri Bal}
\author[1]{Kees Verstoep}
\author[1]{Wan Fokkink}
\author[3]{Jeff Templon}

@@

\begin{abstract}
DIRAC (Distributed Infrastructure with Remote Agent Control) is the Grid
solution designed to support production activities as well as user data analysis
for the Large Hadron Collider beauty experiment. It consists of cooperating
distributed services and a plethora of light-weight agents delivering the
workload to the Grid resources.  Services accept requests from agents and
running jobs, while agents actively fulfill specific goals. Services maintain
database back-ends to store dynamic state information of entities such as jobs,
queues, or staging requests. Agents continuously check for changes in the
service states, and react to these accordingly. The logic of each agent is
rather simple; the main source of complexity lies in their cooperation. These
agents run concurrently, and communicate using the services' databases as a
shared memory for synchronizing the state transitions. Despite the effort
invested in making DIRAC reliable, entities occasionally get into inconsistent
states. Tracing and fixing such behaviors is difficult, given the inherent
parallelism among the distributed components, and the size of the
implementation.

@@

The processing begins with the \textit{Request Preparation Agent}. It selects all the
``New'' entries for files, checks whether they are registered in a special logical
file catalog, and retrieves their metadata. In case of problematic catalog
entries, it can update the status of the CacheReplicas and the related Tasks
entries to ``Failed''. For non-problematic files it carries out transition to
``Waiting''. The \textit{StageRequest Agent} is responsible for placing the actual staging
requests for all ``Waiting'' entries, via dedicated storage middleware that
communicates with the tape backends. These requests are grouped by SE prior to
submission, and carry information about the requested (pin) lifetime of the
replicas to be cached. If certain pathologies are discovered (i.e., lost,
unavailable or zero-sized files on tape), it can update the corresponding
entries to ``Failed'' in a similar manner. Otherwise, they are promoted to
``StageSubmitted''. The agent responsible for monitoring the status of submitted
requests is the \textit{StageMonitor Agent}. It achieves this by interrogating the
storage middleware to see if the ``StageSubmitted'' files are successfully
replicated on disk cache. In case of success, the CacheReplicas and their
corresponding Tasks entries are updated to ``Staged''. Various circumstances of
tape or middleware misbehavior can also fail the staging requests. The last one
in the chain is the \textit{RequestFinalization Agent}. The tasks which are in their
final states (``Staged'' or ``Failed'') are cleared from the database, and callbacks
are performed to the WMS, which effectively wakes up the corresponding jobs. If
there are no more associated tasks for particular replicas, the respective
CacheReplicas entries are also removed.

@@

Given the complexity of the reccurent invalid state transitions of entities within DIRAC, 
we considered the possible race conditions caused by multiple agents updating 
the service states to be the target of our analysis. We limit the scope
to analysis of the following entities: \textit{Tasks} (SMS), \textit{CacheReplicas} (SMS) and \textit{Jobs} (WMS).
This determines our slicing criteria. In the following, we use the SMS as a case study
for describing the established modeling guidelines. They can be 
diractly applied to other DIRAC systems.

As already explained, all agents repeat the same logic in every subsequent iteration:first they
read some entries of interest from the service database, then they process the cached data, and finally
they may write or update entries back to the central database, based on decisions from the processing step.
The frontend for all interractions with the storage are the services, essentially acting as wrappers around it.
Take as an example the code snippet\footnote{This method is $\sim$100 lines of code in reality} 
in Listing~\ref{listing1}, 
from the simplest SMS agent: the \textit{Stage Request Agent}. 
Although much of the code is omitted for clarity, care is taken to preserve
the parts necessary for ilustrating the basic idea.
 (and yet it looks baffling!)

@@

\begin{lstlisting}[float=tp,escapechar=!,basicstyle=\ttfamily\fontsize{7}{7}\selectfont]
  def prepareNewReplicas( self ):
    """ This is the first logical task to be executed 
    """ and manages the New->Waiting transition of the Replicas
    res = self.__getNewReplicas()
    !\colorbox{light-gray}{res = self.getNewReplicas()}!
    if not res['Value']:
      gLogger.info('There were no New replicas found')
      return res
    [...]
    # Obtain the replicas from the FileCatalog
    res = self.__getFileReplicas( fileSizes.keys() )
    if not res['OK']:
      return res
    failed.update( res['Value']['Failed'] )
    terminal = res['Value']['ZeroReplicas']
    fileReplicas = res['Value']['Replicas']
    [...]
    # Check the replicas exist at the requested site
    replicaMetadata = []
    for lfn, requestedSEs in replicas.items():
      lfnReplicas = fileReplicas[lfn]
      for requestedSE, replicaID in requestedSEs.items():
        if not requestedSE in lfnReplicas.keys():
          terminalReplicaIDs[replicaID] = "LFN not registered at requested SE"
          replicas[lfn].pop(requestedSE)
        else:
          replicaMetadata.append((replicaID, lfnReplicas[requestedSE], fileSizes[lfn]))

@@

>>>>

\usepackage{color}
\usepackage{url}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{amsmath}
\usepackage{listings}
\definecolor{light-gray}{gray}{0.80}
%\usepackage[center]{caption}
\DeclareMathSizes{7}{7}{7}{7}
\usepackage{authblk}
\begin{document}
\title{Behavioral Analysis of the Agent-Based Community Grid Solution for the Large Hadron Collider beauty Experiment}
\author[1,3]{Daniela Remenska}
\author[2]{Tim A.C. Willemse}
\author[1]{Henri Bal}
\author[1]{Kees Verstoep}
\author[1]{Wan Fokkink}
\author[3]{Jeff Templon}

@@

\begin{abstract}
DIRAC (Distributed Infrastructure with Remote Agent Control) is the Grid
solution designed to support production activities as well as user data analysis
for the Large Hadron Collider beauty experiment. It consists of cooperating
distributed services and a plethora of light-weight agents delivering the
workload to the Grid resources.  Services accept requests from agents and
running jobs, while agents actively fulfill specific goals. Services maintain
database back-ends to store dynamic state information of entities such as jobs,
queues, or staging requests. Agents continuously check for changes in the
service states, and react to these accordingly. The logic of each agent is
rather simple; the main source of complexity lies in their cooperation. These
agents run concurrently, and communicate using the services' databases as a
shared memory for synchronizing the state transitions. Despite the effort
invested in making DIRAC reliable, entities occasionally get into inconsistent
states. Tracing and fixing such behaviors is difficult, given the inherent
parallelism among the distributed components and the size of the
implementation.

@@

The processing begins with the \textit{Request Preparation Agent}. It selects all the
``New'' entries for files, checks whether they are registered in a special Logical
File Catalog (LFC), and retrieves their metadata. In case of problematic catalog
entries, it can update the status of the CacheReplicas and the related Tasks
entries to ``Failed''. For non-problematic files it carries out transition to
``Waiting''. The \textit{StageRequest Agent} is responsible for placing the actual staging
requests for all ``Waiting'' entries, via dedicated storage middleware that
communicates with the tape backends. These requests are grouped by SE prior to
submission, and carry information about the requested (pin) lifetime of the
replicas to be cached. If certain pathologies are discovered (i.e., lost,
unavailable or zero-sized files on tape), it can update the corresponding
entries to ``Failed'' in a similar manner. Otherwise, they are promoted to
``StageSubmitted''. The agent responsible for monitoring the status of submitted
requests is the \textit{StageMonitor Agent}. It achieves this by interrogating the
storage middleware to see if the ``StageSubmitted'' files are successfully
replicated on disk cache. In case of success, the CacheReplicas and their
corresponding Tasks entries are updated to ``Staged''. Various circumstances of
tape or middleware misbehavior can also fail the staging requests. The last one
in the chain is the \textit{RequestFinalization Agent}. The tasks which are in their
final states (``Staged'' or ``Failed'') are cleared from the database, and callbacks
are performed to the WMS, which effectively wakes up the corresponding jobs. If
there are no more associated tasks for particular replicas, the respective
CacheReplicas entries are also removed.

@@

Given the reccurent invalid state transitions of entities within DIRAC, 
we considered the possible race conditions caused by multiple agents updating 
the service states to be the target of our analysis. We limit the scope
to analysis of the following entities: \textit{Tasks} (SMS), \textit{CacheReplicas} (SMS) and \textit{Jobs} (WMS).
This determines our slicing criteria. In the following, we use the SMS as a case study
for describing the established modeling guidelines. They can be 
directly applied to other DIRAC systems. First, the 
guidelines are informaly described, followed by fragments 
of the mCRL2 model to illustrate the translation.

\subsubsection{Control Abstractions}
As already explained, all agents repeat the same logic in every subsequent iteration:first they
read some entries of interest from the service database, then they process the cached data, and finally
they may write or update entries back, based on decisions from the processing step.
They can be naturally modelled as \textit{recursive processes}.
Take as an example the code snippet\footnote{This method is $\sim$100 lines of code in reality} 
in Listing~\ref{listing1}, 
from the simplest agent: the \textit{Stage Request Agent}. 
Although much of the code is omitted for clarity, the necessary parts for
ilustrating the basic idea are kept.
The first highlighted statement is the selection of all "New'' CacheReplicas
entries. What follows is retrieval of their metadata from the LFC,
external to this subsystem. Subsequently, list and dictionary manipulations
are done to group the retrieved data depending on the outcome.
Two lists of replica IDs are built before the last step: one for the problematic catalog entries,
and one for the sucesful sanity checks.
Finally, the last two highlighted code segments update the states 
of the corresponding CacheReplicas to "Failed'' and ``Waiting'' respectfully.

@@

\begin{lstlisting}[float=tp,escapechar=!,basicstyle=\ttfamily\fontsize{7}{7}\selectfont]
  def prepareNewReplicas( self ):
    """ This is the first logical task to be executed 
    """ and manages the New->Waiting transition of the Replicas
    !\colorbox{light-gray}{res = self.getNewReplicas()}!
    if not res['Value']:
      gLogger.info('There were no New replicas found')
      return res
    [...]
    # Obtain the replicas from the FileCatalog
    res = self.__getFileReplicas( fileSizes.keys() )
    if not res['OK']:
      return res
    failed.update( res['Value']['Failed'] )
    terminal = res['Value']['ZeroReplicas']
    fileReplicas = res['Value']['Replicas']
    [...]
    replicaMetadata = []
    for lfn, requestedSEs in replicas.items():
      lfnReplicas = fileReplicas[lfn]
      for requestedSE, replicaID in requestedSEs.items():
        if not requestedSE in lfnReplicas.keys():
          terminalReplicaIDs[replicaID] = "LFN not registered at requested SE"
          replicas[lfn].pop(requestedSE)
        else:
          replicaMetadata.append((replicaID, lfnReplicas[requestedSE], fileSizes[lfn]))

@@

The logging statements, although critical for operational matters, will
not affect the entries' states, and can be translated to \textit{silent steps} in mCRL2.
Furthermore, instead of tracing back and modeling all variables on which 
the two final lists depend, we can use nondeterminism. 
It is not known upfront which branch execution will follow
for a particular replica, as it depends on external behaviour (i.e. the interactions of the system with its environment)
However, by stubbing-out the communication with the LFC and
most of the local variable manipulations 
that follow, and replacing them with a \textit{nondeterministic choice} between the two
ultimate state updates, we can include both possibilities in the model behavior,
and still preserve correctness. Of course, depending on the context, some 
variable values cannot be ignored, in which case decisivness can be preserved
using \textit{if-then-else} mCRL2 statements. All relevant selection and update 
statements are translated into \textit{parametrized actions}.

\subsubsection{Data Abstractions}
The CacheReplicas entity contains more information besides the state.
Every database entry has a unique identifier, descriptive data such as 
the SE where it resides, its full path, checksum, timestamps etc. 
Model checking can only be performed on closed models, where the domains
of all variables are finite. Since we are only interested in state transitions,
we can collapse most of this descriptive data, and represent this entity as a 
\textit{custom sort} in mCRL2:
%\begin{lstlisting}[tabsize=10,basicstyle=\ttfamily\fontsize{8}{7}\selectfont]
\begin{displaymath}
\begin{align*}
sort\ CacheReplicas = \textbf{struct}\ Start?is\_start\ | \\
			  New?is\_new\ |  \\
		    Waiting?is\_waiting\ | \\
          StageSubmitted?is\_stageSubmitted\ | \\ 
		      Staged?is\_staged\ | \\
		      Failed?is\_failed\ | \\
		  Deleted?is\_deleted\ ;
\end{align*}
\end{displaymath}
%\end{lstlisting}
This defines an enumerated data type with all possible states, together with constructor and recognizer functions.
The Tasks entity is modeled in the same manner. Lists of these sorts can be easily modeled 
in mCRL2 as \begin{math}List(Tasks) \end{math} and 
\begin{math}List(CacheReplicas) \end{math}.
To define the many-to-many relationship between Tasks and CacheReplicas,
we join these data elements in a tuple:
%\begin{lstlisting}[tabsize=10,basicstyle=\ttfamily\fontsize{8}{7}\selectfont]
\begin{displaymath}sort \ Tuple = \textbf{struct}\ p(t:Nat,r:Nat,link:Bool)
\end{displaymath}
%\end{lstlisting}
The first two elements are the list positions of the Tasks and CacheReplicas
entries, while the last one indicates whether a relation between
them exists at a given moment of the system execution. The projections 
($t:Tuple\mapsto Nat$, $r:Tuple\mapsto Nat$, $link:Tuple\mapsto Bool$)
of the tuple on the individual data types are automatically defined with the above
statement.

In reality agents operate on lists of IDs corresponding to the database
entries, so functions for transforming items of type \begin{math}List(CacheReplicas)\end{math}
and \begin{math}List(Tasks)\end{math}
to a list of identifiers (i.e., positions) 
\begin{math}List(Nat)\end{math} are necessary.
One such \textit{data transformation} used in the model is the 
$t2id:List(Tasks)\times Tasks\mapsto List(Nat)$ function, which,
given an existing list of Tasks and a specific Tasks value, returns the list positions
 matching the value.
For example:
% example t2id([one,two,two,three],two)->[1,2] 
\begin{displaymath}
t2id([New,Staged,Staged,Failed],Staged)\rightarrow [1,2] 
\end{displaymath}
Another example is the 
$id2cr:List(Nat)\times Lsit(CacheReplicas)\times CacheReplicas\mapsto List(CacheReplicas)$                                                                                                
function, which can be used to update certain CacheReplicas list entries with a new value, in the following way:
\begin{displaymath}
id2cr([0,1],[Waiting,Staged,New],Failed) \\
 \rightarrow \\
\end{displaymath}
\begin{displaymath}
[Failed,Failed,New]
\end{displaymath}
These data transformations provide a natural way of modeling
the actual database operations.

The mCRL2 language does not allow global variables (or a similar construct).
Therefore, the shared database is modeled as a parametrized wrapper process
that keeps the entries in its local memory, as illustrated in Figure~\ref{fig:CRMemory}.
\begin{figure*}[!t]

$proc\ CacheReplicaMem(d:List(CacheReplicas)) = $

$\sum_{t:CacheReplicas} RPAgent\_selectCacheReplicas\_(cr2id(d,t),t).
CacheReplicaMem(d) +$

$\sum_{l:List(Nat),t:CacheReplicas} RPAgent\_prepareNewReplicas\_(l,t).CacheReplicaMem(id2cr(l,d,t))+$

$\dots\ +$

$\sum_{l:List(Nat),t:CacheReplicas} RFAgent\_removeReplicas\_(l,t).CacheReplicaMem(id2cr(l,d,t))+$

$CacheReplicaMem(d);$

\caption{CacheReplicas memory process}
\label{fig:CRMemory}
\end{figure*}
By means of process communication, the shared list of entries can be obtained
and changed by the agent processes. The
recursive process continously listens and responds
to requests from other processes (agents).
The summation operator allows the process to 
accept any value of the \begin{math}CacheReplicas\end{math}
and \begin{math}List(Nat)\end{math} sort passed by agents.
To ensure that such synchronous communication between the memory process and the agents
is possible, the mCRL2 \begin{math}allow\end{math}
and \begin{math}comm\end{math}
constructs are used
in the \begin{math}init\end{math}
part of the specification:

$ \Delta_{\{RPAgent\_selectCacheReplicas,
    RPAgent\_prepareNewReplicas, } $
  $    _{\dots\\, RFAgent\_removeReplicas\}} $

$ \Gamma_{\{\_RPAgent\_selectCacheReplicas|
   RPAgent\_selectCacheReplicas\_} $
  $  _{\rightarrow RPAgent\_selectCacheReplicas, } $
  $ _{\_RPAgent\_prepareNewReplicas|RPAgent\_prepareNewReplicas\_} $
  $  _{\rightarrow RPAgent\_prepareNewReplicas,\dots\\, } $
 $ _{\_RFAgent\_removeReplicas|RFAgent\_removeReplicas\_} $
  $  _{\rightarrow RFAgent\_removeReplicas\}} $
$ \\ $

To complete the picture, we give an outline of the model for the \begin{math}Request Preparation Agent\end{math}:

$ _{RPAgent = }$
	$ _{\sum_{cc:List(Nat)} \_RPAgent\_selectCacheReplicas(cc,New). }$

	$ _{((cc\neq[])\rightarrow \ (\_RPAgent\_prepareNewReplicas(cc,Failed). } $

	$_{	\sum _{tt:List(Nat)} \_RPAgent\_selectTaskReplicas(tt,cc).  }$

	$_{	((tt\neq[])\rightarrow \_RPAgent\_prepareNewReplicasT(tt,tFailed)\ \diamond \ \tau)\ + }$

	$ _{\_RPAgent\_prepareNewReplicas(cc,Waiting))\ \diamond \ \tau) } $

$ _{.RPAgent; }$
$ \\ $

%Since the services are the frontend for all interractions with the storage,
%they essentially act as wrappers around the storage. By already modeling
%the storage as a recursive process with memory, there is no need to model them explicitely.

Thus, so far we have established: a) a finite domain of abstract values, b) an
abstraction for mapping concrete program values to abstract ones and
c) collection of processes built from parametrized actions describing 
the behavior of the program components.
Substituting concrete operations on real data of interest
with the corresponding operations of the abstract interpretation yields 
our abstract model. Finally, the model is put together as a parallel composition of all
communicating processes.
